Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
Replace non-ascii character
Add url commands

The review process was fairly standard. First, reviewers bid for papers. Then an
automated system matched the bids to determine assignments. After the reviewers
completed their assigned reviews (with an average of 4.9 reviews per paper and
4.1 reviews per reviewer), they used EasyChair\footnote{http://easychair.org/}
to record scores on relevance and comments. Finally, the organizers accessed the
information to decide which papers to associate with the workshop and provided
authors with the comments to help them improve their papers.

@@

\begin{comment}
(Kaitlin didn't really say why? Were the tools used? Did they immediately
bit-rot? Was their development unsustained?)-should ask question about this. Is
there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like
there is: \url{tinyurl.com/qdodbo7}.
%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom (UK). The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point 
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

Chue Hong drew attention to the issue that in the cyber-infrastructure and
high-performance community, hundreds of thousands
of researchers developing software are all too often disregarded or considered the long tail. 
Actually, the numbers point to the fact that they are the mainstream. %and we
%(referring to people including the audience of the talk) are the elite minority.
He emphasized that software is no longer special; it is both essential to and
common in scientific research. A 2014
survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}
conducted by the SSI polled researchers from 15 research-intensive UK
universities (406 respondents covering a representative range of funders,
disciplines, and seniority). The survey reported that 92\% respondents confirmed the use of research
software and 89\% affirmed that it would be impossible or difficult to conduct
research without software. Nevertheless the British research community is just
starting to understand the magnitude of the issue. Whilst many researchers make
use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
SSI survey shows that over half (56\%) of respondents developed their own
research software (which equates to over 140,000 researchers if extrapolated
across the UK) and yet 71\% of all UK researchers had no formal software-development training, having
to rely on their own coding skills.

@@

Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable. Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of "the community"? Some communities have had significant initiatives to improve software discovery, e.g. the NIH Software Discovery Index\footnote{NIH Software Discovery Index: http://softwarediscoveryindex.org/}. Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.

@@

Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website. Using code repositories like GitHub\footnote{GitHub: http://github.com/} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: http://figshare.com/} or Zenodo\footnote{Zenodo: http://zenodo.org/}. Docker\footnote{Docker: htp://www.docker.com/} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.

@@

This group aimed to come up with a categorisation of journals which published software papers. Starting from the list of journals\footnote{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers. These were the Journal of Open Research Software\footnote{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}, PLoS ONE\footnote{http://www.plosone.org/static/guidelines\#software}, Journal of Statistical Software\footnote{http://www.jstatsoft.org/instructions}. Methods in Ecology and Evolution\footnote{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}, Transactions of Mathematical Software\footnote{http://toms.acm.org/Authors.html}, GigaScience\footnote{http://www.gigasciencejournal.com/about}, and PLoS Computational Biology\footnote{http://www.ploscompbiol.org/static/guidelines\#software}.

@@

Journals should provide guidelines about what to consider when reviewing a software submission. A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{http://openresearchsoftware.metajnl.com/about/editorialPolicies}. Journals might also learn from organisations such as the Apache Foundation as to what is good practice for software submissions. Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.

@@

The group discussed how they learned testing.  One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference. The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.  Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.  The difference in his work is that he now knows about frameworks that let him do that systematically.  He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
most of those initial checks were ephemeral.  They live long enough to give the author of the test confidence in the code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.

@@

\begin{table*}[t]
\centering
\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
  \begin{scriptsize}
  \begin{tabular}{l|l|r|r}
 \hline
    Author  &   Tweet  & Retweets &  Favorites
\\ \hline 
% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
%
% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
%\\     &  management plan used by EPSRC software grants    &    &  
% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    & 
%
 Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
\\   &   http://www.software.ac.uk/resources/guides/software-management-plans &    & 
\\   &   we provide a software evaluation tool:   &    & 
\\   &   http://www.software.ac.uk/online-sustainability-evaluation  &    & 
%
\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
\\ & UK researchers write research software w$/$out any formal training &    & 
%
\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
%
\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
%
\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
\\ &   [include] how many faces are up on our community page''   &    & 
%
\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
\\ &  software http://arxiv.org/abs/1407.5117  Work with @arfon  &    & 
% 
%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
% 
\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
%
\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
\\ &   http://www.slideshare.net/kaythaney/     & & 
\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
%
\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & & 
%
%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
%
\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
\\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
\\ & http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/  & & 
%
\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
\\ & Bioinformatics Manifesto: https://github.com/pjotrp/bioinformatics   & 
%
\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3 
\\ & but because influential people are averse to change...''  C. Sunstein     & & 
% 
\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
\\   &  Research Software http://arxiv.org/abs/1407.5981  & & 
%
\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
%
\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
\\ &   $+$ sustainability. Tune in! Live stream: http://ustre.am/17ddh & & 
%
\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
\\ & http://wssspe.researchcomputing.org.uk/wssspe2/agenda/   &  &
\\ & URL for live stream of keynotes \& lightning talks: http://ustre.am/17ddh   &  &
\\ \hline
    \end{tabular}
    \end{scriptsize}
\end{table*}

>>>>

The review process was fairly standard. First, reviewers bid for papers. Then an
automated system matched the bids to determine assignments. After the reviewers
completed their assigned reviews (with an average of 4.9 reviews per paper and
4.1 reviews per reviewer), they used EasyChair\footnote{\url{http://easychair.org/}}
to record scores on relevance and comments. Finally, the organizers accessed the
information to decide which papers to associate with the workshop and provided
authors with the comments to help them improve their papers.

@@

\begin{comment}
(Kaitlin didn't really say why? Were the tools used? Did they immediately
bit-rot? Was their development unsustained?)-should ask question about this. Is
there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like
there is: \url{http://tinyurl.com/qdodbo7}.
%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom (UK). The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point 
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{\url{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

Chue Hong drew attention to the issue that in the cyber-infrastructure and
high-performance community, hundreds of thousands
of researchers developing software are all too often disregarded or considered the long tail. 
Actually, the numbers point to the fact that they are the mainstream. %and we
%(referring to people including the audience of the talk) are the elite minority.
He emphasized that software is no longer special; it is both essential to and
common in scientific research. A 2014
survey\footnote{\url{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}}
conducted by the SSI polled researchers from 15 research-intensive UK
universities (406 respondents covering a representative range of funders,
disciplines, and seniority). The survey reported that 92\% respondents confirmed the use of research
software and 89\% affirmed that it would be impossible or difficult to conduct
research without software. Nevertheless the British research community is just
starting to understand the magnitude of the issue. Whilst many researchers make
use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
SSI survey shows that over half (56\%) of respondents developed their own
research software (which equates to over 140,000 researchers if extrapolated
across the UK) and yet 71\% of all UK researchers had no formal software-development training, having
to rely on their own coding skills.

@@

Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable. Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of "the community"? Some communities have had significant initiatives to improve software discovery, e.g. the NIH Software Discovery Index\footnote{NIH Software Discovery Index: \url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.

@@

Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website. Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}. Docker\footnote{Docker: \url{htp://www.docker.com/}} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.

@@

This group aimed to come up with a categorisation of journals which published software papers. Starting from the list of journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers. These were the Journal of Open Research Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}}, PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}}, Journal of Statistical Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in Ecology and Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}}, Transactions of Mathematical Software\footnote{\url{http://toms.acm.org/Authors.html}}, GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS Computational Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.

@@

Journals should provide guidelines about what to consider when reviewing a software submission. A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}. Journals might also learn from organisations such as the Apache Foundation as to what is good practice for software submissions. Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.

@@

The group discussed how they learned testing.  One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference. The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.  Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.  The difference in his work is that he now knows about frameworks that let him do that systematically.  He realized he was always building on top of other people's software, and eventually realized that he should adopt those good software habits.  He also commented that
most of those initial checks were ephemeral.  They live long enough to give the author of the test confidence in the code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.

@@

\begin{table*}[t]
\centering
\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
  \begin{scriptsize}
  \begin{tabular}{l|l|r|r}
 \hline
    Author  &   Tweet  & Retweets &  Favorites
\\ \hline 
% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
%
% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
%\\     &  management plan used by EPSRC software grants    &    &  
% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    & 
%
 Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    & 
\\   &   we provide a software evaluation tool:   &    & 
\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    & 
%
\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
\\ & UK researchers write research software w$/$out any formal training &    & 
%
\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
%
\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
%
\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
\\ &   [include] how many faces are up on our community page''   &    & 
%
\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    & 
% 
%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
% 
\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
%
\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
\\ &   \url{http://www.slideshare.net/kaythaney/}     & & 
\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
%
\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & & 
%
%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
%
\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
\\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & & 
%
\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   & 
%
\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3 
\\ & but because influential people are averse to change...''  C. Sunstein     & & 
% 
\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & & 
%
\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
%
\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & & 
%
\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
\\ \hline
    \end{tabular}
    \end{scriptsize}
\end{table*}
